---
title: "Is The Seven-Year Itch Real?"
subtitle: "Assignment 4 ETC5512"
author: "Nattamon Wattanapenpaiboon"
format: html
editor: visual
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->
## Data and Documenting

### Topic of this work?

This blog post investigates whether the cultural idea of the *Seven‑Year Itch*—the belief that marriages tend to break down around the seven‑year mark—has any empirical support in global divorce data.

To explore this, I also consider several sub‑questions:

- How do divorce counts vary across different marriage‑duration categories

- Do countries show similar or different patterns in when marriages tend to end?

- Has the timing of divorce peaks shifted over the past 30 years?


The Seven‑Year Itch is widely referenced in popular culture but rarely tested with global, longitudinal data. The dataset provides a unique opportunity to examine a familiar idea through a demographic lens.

### Data description
This analysis uses the “Divorces by duration of marriage” dataset from the Demographic Statistics Database maintained by the United Nations Statistics Division (UNSD). The data form part of the United Nations Demographic Yearbook system, which collects official demographic and social statistics from national statistical offices through annual questionnaires. Further information is described below. Only the key data‑cleaning decisions that affect interpretation are summarised in this blog post. Full details of the wrangling process, including variable transformations, are documented in the accompanying README.txt and data‑dictionary.csv files.

#### Metadata
Dataset name: Divorces by duration of marriage

Data provider: United Nations Statistics Division (UNSD)

Temporal coverage: Collected since 1948 — analysis focuses on the most recent 30 years (1995–2025)

Spatial coverage: 105 countries or areas (questionnaires sent to over 230 countries and areas) — analysis focuses on European countries

Licence / Terms of use: Released under the UNdata Terms of Use, allowing free reuse with attribution but prohibiting endorsement and commercial misuse

Format: CSV

Last updated: 26 February 2026

Data type: Administrative population data derived from national civil registration systems. Although census‑style in that reporting countries provide complete counts of all recorded divorces, the global dataset is incomplete because some countries do not legally recognise divorce or do not submit divorce statistics to the UN Demographic Yearbook.

Privacy / Ethics: Aggregated data with no direct identifiers

#### Key Limitations
- Incomplete global coverage — Many countries do not submit divorce‑duration statistics to the UN, resulting in substantial geographic gaps. Europe provides the most complete and consistent divorce‑duration data, so the analysis focuses on European countries to ensure comparability and minimise bias introduced by missing or irregular data

- Differences in duration‑of‑marriage categories — Countries use different binning schemes, requiring standardisation and sometimes forcing loss of detail.

- Legal and institutional differences — Countries vary widely in divorce laws, waiting periods, and registration systems. Some report divorces by year of occurrence while others use year of registration, affecting comparability across countries.

### Data Download
The following section describes how the data can be downloaded from its source and what initial data wrangling steps have been taken.
1. Access the dataset via <https://data.un.org/> 
2. Search for 'Divorces by duration of marriage' or access it directly via <https://data.un.org/Data.aspx?q=Divorces+by+duration+of+marriage&d=POP&f=tableCode%3a87>
3. Select "Download"
```{r}
# Data cleaning before analysis

# ============================================================
# 1. Load packages and raw data
# ============================================================

library(tidyverse)
library(janitor)
library(countrycode)
library(stringr)
library(sf)
library(rnaturalearthdata)
library(rnaturalearth)

divorce_data <- read_csv("data/UNdata_Export_20260528_021328529.csv")

# ============================================================
# 2. Initial cleaning (remove footnotes)
# ============================================================
divorce_data <- divorce_data |>
  clean_names() |>
  mutate(
    duration_of_marriage = str_trim(duration_of_marriage),
    year = str_trim(year)
  ) |>
  # Keep only valid 4-digit years (removes footnotes)
  filter(str_detect(year, "^[0-9]{4}$"))

# ============================================================
# 3. Construct percent-of-total (create comparable measures)
# ============================================================
# Step 1: Extract and rank "Total" rows by data quality
totals <- divorce_data |>
  filter(duration_of_marriage == "Total") |>
  mutate(
    quality_rank = case_when(
      reliability == "Final figure, complete" ~ 1,
      record_type == "Data tabulated by year of occurrence" ~ 2,
      record_type == "Data tabulated by year of registration" ~ 3,
      TRUE ~ 4
    )
  ) |>
  arrange(country_or_area, year, quality_rank) |>
  group_by(country_or_area, year) |>
  slice(1) |>   # <-- Keep only the best single row
  ungroup() |>
  select(country_or_area, year, total_value = value)


# Step 2: Keep duration categories only
details <- divorce_data |>
  filter(!duration_of_marriage %in% c("Total", "Not stated"))


# Step 3: Join totals back in and compute percent_of_total
clean_divorce <- details |>
  left_join(totals, by = c("country_or_area", "year")) |>
  mutate(
    percent_of_total = value / total_value,
    # Add column continent for further grouping analysis
    continent = countrycode(country_or_area, "country.name", "continent")
  ) |>
  # Keep only recent 30 years
  filter(as.numeric(year) >= 1995)

# ============================================================
# 4. Standardise duration-of-marriage categories
# ============================================================
clean_divorce <- clean_divorce |>
  mutate(
    # Extract all numbers from the bucket (handles "1", "1-3", "10 +", "Less than 5")
    nums = str_extract_all(duration_of_marriage, "\\d+"),
    nums = lapply(nums, as.numeric),
    
    # Determine the maximum number in the bucket
    max_years = sapply(nums, function(x) if(length(x) == 0) NA else max(x)),
    
     # Adjust for "Less than X" so upper bound becomes X - 1
    max_years = case_when(
      str_detect(str_to_lower(duration_of_marriage), "less") ~ max_years - 1,
      TRUE ~ max_years
    ),
    # Standardise into global bins
    duration_std = case_when(
      max_years <= 1 ~ "0-1",
      max_years <= 2 ~ "1-2",
      max_years <= 4 ~ "3-4",
      max_years <= 9 ~ "5-9",
      max_years >= 10 ~ "10+",
      TRUE ~ NA_character_
    )
  ) |>
  select(-nums, -max_years, - duration_of_marriage)

# ============================================================
# 5. Harmonise country names for mapping
# ============================================================
world <- ne_countries(scale = "medium", returnclass = "sf")
world_names <- world |>
  st_drop_geometry() |>
  select(name)
# Find the mismatch in names spelling
mismatch <- clean_divorce |>
  distinct(country_or_area) |>
  anti_join(world_names, by = c("country_or_area" = "name"))


# Build the lookup table
name_fix <- tribble(
  ~country_or_area, ~name,
  "Brunei Darussalam", "Brunei",
  "China, Macao SAR", "Macau",
  "Iran (Islamic Republic of)", "Iran",
  "Netherlands (Kingdom of the)", "Netherlands",
  "Republic of Korea", "South Korea",
  "Republic of Moldova", "Moldova",
  "Russian Federation", "Russia",
  "State of Palestine", "Palestine",
  "Türkiye", "Turkey",
  "United Kingdom of Great Britain and Northern Ireland", "United Kingdom",
  "Venezuela (Bolivarian Republic of)", "Venezuela"
)

# Apply to fixes
clean_divorce <- clean_divorce |>
  left_join(name_fix, by = "country_or_area") |>
  mutate(
    country = if_else(is.na(name), country_or_area, name)
  ) |>
  select(-name, -country_or_area) |>
  relocate(country)

# ============================================================
# 6. Assess global data completeness
# ============================================================
data_coverage <- clean_divorce |>
  group_by(country) |>
  summarise(has_data = any(!is.na(value))) |>
  mutate(has_data = ifelse(has_data, "Data available", "No data"))

map_coverage <- world |>
  left_join(data_coverage, by = c("name" = "country"))

# Europe outline for emphasis
europe_outline <- world |>
  filter(continent == "Europe") |>
  st_combine() |>
  st_as_sf()

# Visualise completeness
ggplot(map_coverage) +
  geom_sf(aes(fill = has_data), color = "white", size = 0.1) +
  geom_sf(data = europe_outline, fill = NA, color = "black", size = 0.8) +
  scale_fill_manual(
    values = c("Data available" = "#4daf4a",   # green
               "No data" = "grey90")
  ) +
  theme_minimal() +
  labs(
    title = "Countries with Available Divorce-Duration Data (UN Demographic Yearbook)",
    fill = "Data Availability"
  )

# ============================================================
# 7. Final dataset preparation (Restrict dataset to Europe)
# ============================================================
# The dataset was filtered to include only European countries, reflecting the region with the most complete reporting.
europe_divorce <- clean_divorce |> filter(continent == "Europe") |> 
  # Keep only relevant columns
  select(-area, -record_type, -reliability, -source_year, -value_footnotes, -continent)

```
#### Key Data Processing Decisions
**Standardising duration-of-marriage categories**
Countries report duration‑of‑marriage categories using inconsistent formats, including single years, multi‑year ranges, open‑ended values (e.g., “10+”), and “Less than X” labels. To enable cross‑country comparison, I extracted all numeric values from each label and used the maximum value as the upper bound of the category. For “Less than X” labels, I assumed the category covers 0 to (X–1) years, as the UN does not specify the exact boundaries. These upper bounds were then collapsed into five global bins (0–1, 1–2, 3–4, 5–9, 10+). This approach standardises heterogeneous reporting but introduces limitations, including loss of granularity, potential misclassification of ambiguous labels, and uncertainty about how “Less than X” categories are defined across countries.

**Restricting the dataset to Europe**
Before analysis, data availability was assessed by identifying whether each country had any non‑missing divorce‑duration observations. This completeness indicator was joined to the world map and visualised. The map showed that Europe has the highest concentration of countries with available data, while many regions outside Europe have sparse or missing reporting.
Given the substantial variation in completeness across regions, the final dataset was restricted to European countries. This decision ensures that the subsequent analysis is based on the region with the most complete and reliable reporting. Only relevant analytical columns were retained in the final dataset.

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### Why explore the Seven-Year Itch?

The *Seven‑Year Itch* is one of those ideas that refuses to die — a cultural belief that marriages tend to break down around the seven‑year mark. It shows up in films, advice columns, and casual conversations, yet rarely gets tested with real data. That gap between myth and measurable reality makes it worth examining.

Understanding *when* marriages tend to end isn’t just trivia. Divorce timing reflects social norms, economic pressures, and how relationships evolve over the life course. If certain durations consistently show higher divorce counts, those patterns can reveal something meaningful about the stresses couples face at different stages.

### Dataset for analysing the duration of marriage before divorce in Europe
The United Nations Statistics Division’s “Divorces by duration of marriage” dataset provides a rare opportunity to examine how long marriages last before ending across countries and over time. However, reporting completeness varies widely around the world. Many regions have sparse or inconsistent data, while Europe stands out as the area with the most comprehensive and continuous reporting of divorce‑duration statistics. To keep the analysis focused on contemporary patterns, this blog uses data from the most recent thirty years (1995–2025).
For this reason, the analysis in **this blog focuses on European countries**, where the data are sufficiently complete to meaningfully explore whether a seven‑year peak in divorce timing actually appears — or whether the cultural myth collapses under empirical evidence.

```{r}
#| echo: false

ggplot(map_coverage) +
  geom_sf(aes(fill = has_data), color = "white", size = 0.1) +
  scale_fill_manual(
    values = c("Data available" = "#4daf4a",
               "No data" = "grey90")
  ) +
  theme_minimal() +
  labs(
    title = "Where Divorce-Duration Data Exists",
    subtitle = "Europe stands out as the region with the most complete reporting",
    fill = NULL
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
```


### What does divorce timing look like across Europe?
The bar graph below shows how common divorce is at different stages of marriage across Europe. The early‑marriage categories are shaded in grey to fade into the background, while the 5–9‑year and 10+‑year durations are highlighted in blue. This colour choice emphasises the mid‑marriage period where the Seven‑Year Itch is said to occur, helping us see whether the data actually supports the myth.
```{r}
avg_duration <- europe_divorce |> 
  mutate(duration_std = factor(duration_std,
                               levels = c("0-1", "1-2", "3-4", "5-9", "10+"))) |>
  group_by(duration_std) |>
  summarise(mean_pct = mean(percent_of_total, na.rm = TRUE))

ggplot(avg_duration, aes(x = duration_std, y = mean_pct, fill = duration_std)) +
  geom_col() +
  scale_fill_manual(
    values = c(
      "0-1" = "#d9d9d9",   # light grey
      "1-2" = "#bdbdbd",   # medium grey
      "3-4" = "#969696",   # darker grey
      "5-9" = "#3182bd",   # strong blue (highlight)
      "10+" = "#08519c"    # darker blue (highlight)
    )
  ) +
  labs(
    title = "Average Divorce Timing Across Europe",
    x = "Years of Marriage",
    y = "Average % of Divorces",
    fill = "Duration"
  ) +
  theme_minimal()

```
Across Europe, divorces are relatively uncommon in the first few years of marriage. The share rises steadily and reaches its highest levels between 3–4 years and 5–9 years of marriage. These mid‑marriage durations account for the largest proportion of divorces, while very early divorces (0–1 or 1–2 years) are much less common. This pattern suggests that marital strain tends to accumulate over time rather than appearing immediately — and that the “Seven‑Year Itch” sits within a broader mid‑marriage peak rather than standing out as a sharp spike.


### Where in Europe do divorces peak?
While the Europe‑wide averages show that divorces tend to cluster in the mid‑marriage years, this pattern is not necessarily identical across all countries. Some nations may experience earlier peaks, while others see divorces occurring later in marriage. To understand these differences, it helps to look at the geography of divorce timing rather than just the overall distribution.

The map below shows the typical divorce duration for each European country, based on **the most common (peak) duration category** observed between 1995 and 2025. Each country is coloured according to the marriage‑length bin that accounts for the largest share of its divorces. Early‑marriage categories (0–4 years) are shaded in grey, while the mid‑marriage and long‑marriage durations (5–9 years and 10+ years) are highlighted in blue to emphasise the period where the “Seven‑Year Itch” is expected to appear.

```{r}
#| echo: false
# Compute each country’s average duration bin
country_peak <- europe_divorce |>
  group_by(country, duration_std) |>
  summarise(avg_pct = mean(percent_of_total, na.rm = TRUE)) |>
  slice_max(avg_pct, n = 1) |>
  ungroup()

# Join country_peak to the map
europe_map <- world |>
  filter(continent == "Europe") |>
  left_join(country_peak, by = c("name" = "country")) |>
  mutate(duration_std = factor(duration_std,
                               levels = c("0-1", "1-2", "3-4", "5-9", "10+")))

# Reproject to a Europe-friendly projection
europe_map_proj <- st_transform(europe_map, 3035)

  ggplot(europe_map_proj) +
  geom_sf(aes(fill = duration_std), color = "white", size = 0.2) +
  scale_fill_manual(
    values = c(
      "0-1" = "#d9d9d9",
      "1-2" = "#bdbdbd",
      "3-4" = "#969696",
      "5-9" = "#3182bd",
      "10+" = "#08519c"
    ),
    na.value = "grey90"
  ) +
  theme_minimal() +
  labs(
    title = "Typical Divorce Timing Across Europe",
    subtitle = "Countries coloured by their most common divorce duration (1995–2025)",
    fill = "Peak Duration"
  )
  
  country_peak_summary <- country_peak |>
  count(duration_std, name = "num_countries") |>
  arrange(duration_std)

country_peak_summary
```
Across Europe, only a small number of countries reach their divorce peak within the first four years of marriage. For most countries, however, the most common divorce durations fall in the 5–9‑year and 10+-year categories, which dominate the map. Early‑marriage divorces do occur—and in a few countries they are the most frequent—but they remain the exception rather than the rule. The broader pattern shows that many European marriages tend to last several years before breaking down, suggesting that marital strain typically accumulates over time rather than emerging immediately.

This geographic pattern reinforces the earlier bar chart: Europe as a whole shows a broad mid‑marriage rise in divorces, and the map demonstrates that this tendency is shared across most countries. While there is some variation in how early or late the peak occurs, the overall picture is consistent—mid‑ and long‑marriage divorces are far more common than early ones.

### Has divorce timing shifted over the last 30 Years?

Now that we’ve seen the overall European pattern and the geographic differences between countries, the next question is whether divorce timing has changed over the last three decades. Social norms, economic conditions, and family structures have all evolved since the mid‑1990s, and these shifts may influence when marriages are most likely to end. To explore this, we track how the share of divorces in each duration category has changed from 1995 to 2025.
```{r}
# Prepare data for time trend
trend_data <- europe_divorce |>
  mutate(
    year = as.numeric(year),
    duration_std = factor(
      duration_std,
      levels = c("0-1", "1-2", "3-4", "5-9", "10+")
    )
  ) |>
  group_by(year, duration_std) |>
  summarise(
    mean_pct = mean(percent_of_total, na.rm = TRUE),
    .groups = "drop"
  )

# Plot: stacked area chart
ggplot(trend_data, aes(x = year, y = mean_pct, fill = duration_std)) +
  geom_area(alpha = 0.9) +
  scale_fill_manual(
    values = c(
      "0-1" = "#d9d9d9",
      "1-2" = "#bdbdbd",
      "3-4" = "#969696",
      "5-9" = "#3182bd",
      "10+" = "#08519c"
    )
  ) +
  labs(
    title = "How Divorce Timing Has Shifted Over Time",
    subtitle = "Average share of divorces in each duration category, 1995–2025",
    x = "Year",
    y = "Average % of Divorces",
    fill = "Duration"
  ) +
  theme_minimal()


```
A clear pattern emerges from the time‑trend data. Early‑marriage divorces (0–4 years) remain consistently low throughout the entire period, showing no sign of becoming more common. In contrast, divorces after ten or more years of marriage have gradually increased, becoming the dominant category across Europe. The 5–9‑year group remains stable as the second‑largest category, reinforcing the idea of a broad mid‑marriage peak rather than a sharp spike at seven years. Taken together, these trends suggest that European marriages are lasting longer before breaking down, and that the “Seven‑Year Itch” is not intensifying over time. Instead, long‑marriage divorces have become increasingly prominent, pointing to deeper structural and social changes rather than a fixed psychological milestone.

### So does the Seven-Year Itch exist?

Taken together, the evidence shows that the “Seven‑Year Itch” is more cultural shorthand than statistical reality. Across Europe, divorces rarely peak early in marriage, and the most common break‑up point consistently falls well beyond the seven‑year mark. Geographic patterns reveal broad agreement across countries, and the long‑term trends show that divorces are, if anything, happening later rather than earlier. While mid‑marriage remains a period of heightened vulnerability, there is no sharp spike at seven years—only a gradual rise and a long plateau. In this sense, the Seven‑Year Itch captures a real idea—that relationships evolve and sometimes strain over time—but it does not correspond to a precise moment in the data.


6.  References

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.
:::


