---
title: "Is The Seven-Year Itch Real?"
subtitle: "Assignment 4 ETC5512"
author: "Nattamon Wattanapenpaiboon"
format: html
editor: visual
execute:
  warning: false
  message: false
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data and Documenting

### Topic of this work

This blog post investigates whether the cultural idea of the **Seven-Year Itch**—the belief that marriages tend to break down around the seven-year mark—has any empirical support in global divorce data. Although the concept is widely referenced in popular culture, it has rarely been examined using large-scale longitudinal evidence. By analysing divorce patterns across countries and over time, this dataset provides a unique opportunity to evaluate a familiar cultural belief through a demographic lens.

### Data description

This analysis uses the *Divorces by duration of marriage* dataset from the United Nations Statistics Division (UNSD), which provides official administrative counts of all recorded divorces by duration of marriage. These are census‑style population data derived from national civil registration systems, making them well‑suited to examining when divorces typically occur and whether they cluster around the so‑called “Seven‑Year Itch.” The dataset covers 105 countries, with the most consistent reporting for European countries. For this analysis, the period 1995-2025 is used to provide a clean and comparable 30-year window. It is released under the UNdata Terms of Use, permitting academic reuse with attribution, and contains only aggregated counts with no personal identifiers, so there are no privacy risks. Key limitations include incomplete global coverage, variation in duration‑of‑marriage categories, and differences in national divorce laws and registration practices.

Comprehensive metadata and data‑transformation details are documented in the README.txt and data‑dictionary.csv, while full citations appear in the Blog Post tab.

### Data Download

The following section describes how the data can be downloaded from its source and what initial data wrangling steps have been taken.

1\. Access the dataset via [https://data.un.org](https://data.un.org/){.uri}

2\. Search for 'Divorces by duration of marriage' or access it directly via <https://data.un.org/Data.aspx?q=Divorces+by+duration+of+marriage&d=POP&f=tableCode%3a87>

3\. Select "Download"

### Data Cleaning

```{r}
# Data cleaning before analysis

# ============================================================
# 1. Load packages and raw data
# ============================================================

library(tidyverse)
library(janitor)
library(countrycode)
library(stringr)
library(sf)
library(rnaturalearthdata)
library(rnaturalearth)


divorce_data <- read_csv("data/UNdata_Export_20260528_021328529.csv")

# ============================================================
# 2. Initial cleaning (remove footnotes)
# ============================================================
divorce_data <- divorce_data |>
  clean_names() |>
  mutate(
    duration_of_marriage = str_trim(duration_of_marriage),
    year = str_trim(year)
  ) |>
  # Keep only valid 4-digit years (removes footnotes)
  filter(str_detect(year, "^[0-9]{4}$"))

# ============================================================
# 3. Construct percent-of-total (create comparable measures)
# ============================================================
# 3.1 Extract and rank "Total" rows by data quality
totals <- divorce_data |>
  filter(duration_of_marriage == "Total") |>
  mutate(
    quality_rank = case_when(
      reliability == "Final figure, complete" ~ 1,
      record_type == "Data tabulated by year of occurrence" ~ 2,
      record_type == "Data tabulated by year of registration" ~ 3,
      TRUE ~ 4
    )
  ) |>
  arrange(country_or_area, year, quality_rank) |>
  group_by(country_or_area, year) |>
  slice(1) |>   # <-- Keep only the best single row
  ungroup() |>
  select(country_or_area, year, total_value = value)


# 3.2 Keep duration categories only
details <- divorce_data |>
  filter(!duration_of_marriage %in% c("Total", "Not stated"))

# ==========================================================================
# 4. Standardise duration-of-marriage categories BEFORE computing percentages
# ==========================================================================
details_std <- details |>
  mutate(
    # Extract all numbers from the bucket (handles "1", "1-3", "10 +", "Less than 5")
    nums = str_extract_all(duration_of_marriage, "\\d+"),
    nums = lapply(nums, as.numeric),
    
    # Determine the maximum number in the bucket
    max_years = sapply(nums, function(x) if(length(x) == 0) NA else max(x)),
    
     # Adjust for "Less than X" so upper bound becomes X - 1
    max_years = case_when(
      str_detect(str_to_lower(duration_of_marriage), "less") ~ max_years - 1,
      TRUE ~ max_years
    ),
    # Standardise into global bins
    duration_std = case_when(
      max_years <= 1 ~ "0-1",
      max_years <= 2 ~ "1-2",
      max_years <= 4 ~ "3-4",
      max_years <= 9 ~ "5-9",
      max_years >= 10 ~ "10+",
      TRUE ~ NA_character_
    )
  ) |>
  select(-nums, -max_years, - duration_of_marriage)

# ==========================================================================
# 3.3 Join totals back in and compute percent_of_total
clean_divorce <- details_std |>
  left_join(totals, by = c("country_or_area", "year")) |>
  group_by(country_or_area, year, duration_std, total_value) |>
  summarise(
    value = sum(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    percent_of_total = value / total_value,
    
# ==========================================================================
# 5. Add continent variable and filter to analysis period (1995–2025)
# ==========================================================================
    # Add column continent for further grouping analysis
    continent = countrycode(country_or_area, "country.name", "continent"),
  # Keep only recent 30 years
  year = as.numeric(year)) |>
  filter(year >= 1995)

# ============================================================
# 6. Harmonise country names and restrict to only Europe
# ============================================================
world <- ne_countries(scale = "medium", returnclass = "sf")
world_names <- world |>
  st_drop_geometry() |>
  select(name)
# Find the mismatch in names spelling
mismatch <- clean_divorce |>
  distinct(country_or_area) |>
  anti_join(world_names, by = c("country_or_area" = "name"))


# Build the lookup table
name_fix <- tribble(
  ~country_or_area, ~name,
  "Brunei Darussalam", "Brunei",
  "China, Macao SAR", "Macau",
  "Iran (Islamic Republic of)", "Iran",
  "Netherlands (Kingdom of the)", "Netherlands",
  "Republic of Korea", "South Korea",
  "Republic of Moldova", "Moldova",
  "Russian Federation", "Russia",
  "State of Palestine", "Palestine",
  "Türkiye", "Turkey",
  "United Kingdom of Great Britain and Northern Ireland", "United Kingdom",
  "Venezuela (Bolivarian Republic of)", "Venezuela"
)

# Apply to fixes
clean_divorce <- clean_divorce |>
  left_join(name_fix, by = "country_or_area") |>
  mutate(
    country = if_else(is.na(name), country_or_area, name)
  ) |>
  select(-name, -country_or_area) |>
  relocate(country)

# The dataset was filtered to include only European countries, reflecting the region with the most complete reporting.
europe_divorce <- clean_divorce |> filter(continent == "Europe") 

# ============================================================
# 7. Save the cleaned dataset
# ============================================================
write_csv(europe_divorce, "data/europe_divorces_by_duration_of_marriage.csv")

```

```{r}
#| echo: false

# Constructing data dictionary

data_dict <- tibble(
  variable = names(europe_divorce),
  type = sapply(europe_divorce, typeof),
  example = sapply(europe_divorce, function(x) {
    # Handle empty columns safely
    if (length(x) == 0) return(NA)
    # Sample one non-missing value
    sample(x[!is.na(x)], 1)
  })
) |>
  mutate(description = c("Country name.",
                         "Calendar year of the divorce statistics.", 
                         "Standardised duration‑of‑marriage category.",
                         "Total number of divorces for a given country–year.",
                         "Number of divorces in the given duration category for that country–year.",
                         "Share of divorces in the given duration category.",
                         "Continent to which the country belongs."), 
         transformation = c("Harmonised country names to a consistent naming scheme (rnaturalearth R package)",
                              "Converted to numeric. Filtered to retain only valid years between 1995 and 2025.",
                              "Collapsed heterogeneous duration labels into five consistent bins: 0-1, 1-2, 3-4, 5-9, 10+",
                              "Extracted from the UN “Total divorces” series.",
                              "Summed across raw UN rows when multiple original categories collapsed into the same duration bin.",
                              "Computed as value / total_value",
                              "Added using countrycode R package"
               )) |>  relocate(description, .before = example) |>
  relocate(transformation, .before = example)

write_csv(data_dict, "data/data_dictionary.csv")
```

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### Why explore the Seven-Year Itch?

The *Seven‑Year Itch* is one of those ideas that refuses to die — a cultural belief that marriages tend to break down around the seven‑year mark. It shows up in films, advice columns, and casual conversations, yet rarely gets tested with real data. That gap between myth and measurable reality makes it worth examining.

Understanding *when* marriages tend to end isn’t just trivia. Divorce timing reflects social norms, economic pressures, and how relationships evolve over the life course. If certain durations consistently show higher divorce counts, those patterns can reveal something meaningful about the stresses couples face at different stages.

### Dataset for analysing the duration of marriage before divorce in Europe

The United Nations Statistics Division’s “Divorces by duration of marriage” dataset provides a rare opportunity to examine how long marriages last before ending across countries and over time. However, reporting completeness varies widely around the world. Many regions have sparse or inconsistent data, while Europe stands out as the area with the most comprehensive and continuous reporting of divorce‑duration statistics. To keep the analysis focused on contemporary patterns, this blog uses data from the most recent thirty years (1995–2025). For this reason, the analysis in **this blog focuses on European countries**, where the data are sufficiently complete to meaningfully explore whether a seven‑year peak in divorce timing actually appears — or whether the cultural myth collapses under empirical evidence.

```{r}
#| echo: false

# 1. Identify countries with at least one non-missing duration observation
coverage <- clean_divorce |>
  group_by(country) |>
  summarise(has_data = any(!is.na(percent_of_total))) |>
  mutate(has_data = if_else(has_data, "Data available", "No data"))

# 2. Join to world map
map_coverage <- world |>
  left_join(coverage, by = c("name" = "country")) |>
  mutate(
    has_data = replace_na(has_data, "No data")
  )

ggplot(map_coverage) +
  geom_sf(aes(fill = has_data), color = "white", size = 0.1) +
  scale_fill_manual(
    values = c("Data available" = "#4daf4a",
               "No data" = "grey90")
  ) +
  theme_minimal() +
  labs(
    title = "Where Divorce Data Exists",
    subtitle = "Europe stands out as the region with the most complete reporting",
    fill = NULL
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
```

### What does divorce timing look like across Europe?

The bar graph below shows how common divorce is at different stages of marriage across Europe. The colours follow a light‑to‑dark gradient, with lighter shades representing earlier durations and darker shades representing longer ones.

```{r}
#| echo: false
avg_duration <- europe_divorce |>
   mutate(duration_std = factor(
      duration_std,
      levels = c("0-1", "1-2", "3-4", "5-9", "10+"))) |>
  group_by(year, duration_std) |>
  summarise(yearly_avg = mean(percent_of_total, na.rm = TRUE)) |>
  ungroup() |>
  group_by(duration_std) |>
  summarise(mean_pct = mean(yearly_avg, na.rm = TRUE))
  

ggplot(avg_duration, aes(x = duration_std, y = mean_pct, fill = duration_std)) +
  geom_col() +
  scale_fill_manual(
    values = c(
      "0-1" = "#deebf7",   
      "1-2" = "#c6dbef",   
      "3-4" = "#9ecae1",   
      "5-9" = "#6baed6",   
      "10+" = "#3182bd"    
    )
  ) +
  labs(
    title = "Average Divorce Timing Across Europe",
    x = "Years of Marriage",
    y = "Average Share of Divorces",
    fill = "Duration"
  ) +
  theme_minimal()

```

Across Europe, divorces are relatively uncommon in the first few years of marriage. The share rises steadily and reaches its highest levels between 3–4 years and 5–9 years of marriage. These mid‑marriage durations account for the largest proportion of divorces, while very early divorces (0–1 or 1–2 years) are much less common. This pattern suggests that marital strain tends to accumulate over time rather than appearing immediately — and that the “Seven‑Year Itch” sits within a broader mid‑marriage peak rather than standing out as a sharp spike.

### Where in Europe do divorces occur earlier or later in a marriage?

Across Europe, divorces are least common in the very early years of marriage, with only a small share occurring within the first one or two years. The proportion grows gradually as marriages last longer, with noticeably more divorces occurring after several years together. Mid‑duration marriages show a moderate rise, but the largest share of divorces happens among couples who have been married for a decade or more. Overall, the pattern suggests that divorce becomes more likely as marriages accumulate time and complexity, rather than peaking sharply in the early or middle years.

While the Europe‑wide averages show that divorces tend to cluster in the long‑marriage years, this pattern is not necessarily identical across all countries. Some nations may experience earlier peaks, while others see divorces occurring later in marriage. To understand these differences, it helps to look at the geography of divorce timing rather than just the overall distribution.

The map below shows the typical length of marriage before divorce in each European country, based on a **weighted average of divorce‑duration categories from 1995 to 2025**. To summarise the open‑ended “10+ years” category, I assume a midpoint of 15 years, reflecting the fact that many divorces in this group occur well beyond the ten‑year mark. Countries are coloured according to the resulting average number of years couples were married before divorcing, grouped into broad duration categories for clarity.

```{r}
#| echo: false

bucket_mid <- c(
  "0-1" = 0.5,
  "1-2" = 1.5,
  "3-4" = 3.5,
  "5-9" = 7,
  "10+" = 15
)

country_year_avg <- europe_divorce |>
  mutate(mid = bucket_mid[duration_std]) |>
  group_by(country, year) |>
  summarise(
    avg_duration_year = sum(mid * percent_of_total, na.rm = TRUE),
    .groups = "drop"
  )

country_avg <- country_year_avg |>
  group_by(country) |>
  summarise(
    avg_duration = mean(avg_duration_year, na.rm = TRUE),
    .groups = "drop"
  )

# Join country_avg to the map
europe_map <- world |>
  filter(continent == "Europe") |>
  left_join(country_avg, by = c("name" = "country"))|>
  mutate(
    duration_cat = cut(
      avg_duration,
      breaks = c(0.5, 1, 3, 5, 10, Inf),
      labels = c("0-1", "1-2", "3-4", "5-9", "10+"),
      include.lowest = TRUE,
      right = FALSE
    )
  )

# Plot map
ggplot(europe_map) +
  geom_sf(aes(fill = duration_cat), color = "white", size = 0.2) +
  scale_fill_manual(
    values = c(
      "0-1" = "#deebf7",
      "1-2" = "#c6dbef",
      "3-4" = "#9ecae1",
      "5-9" = "#6baed6",
      "10+" = "#3182bd"
    ),
    na.value = "grey90",
    name = "Avg Years"
  ) +
  coord_sf(
    xlim = c(-25, 45),
    ylim = c(34, 72),
    expand = FALSE
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank()
  ) +
  labs(
    title = "Average Duration of Marriage Before Divorce Across Europe",
    subtitle = "Weighted average based on duration categories (1995–2025)"
  )


```

The map reveals a clear continental pattern in how long marriages typically last before divorce. Although every country experiences divorces at a range of durations, their weighted averages show that most European countries tend to divorce later, not earlier. A large share of countries fall into the 10+‑year category once the long‑marriage bucket is weighted appropriately, indicating that divorces commonly occur well after a decade of marriage. Others cluster in the 5–9‑year range, reflecting a mid‑marriage concentration.

This geographic pattern aligns with the broader trend visible in the earlier bar chart: Europe’s divorce timing is shaped primarily by mid‑ and long‑marriage breakdowns, with early‑marriage divorces playing a comparatively small role. While there is meaningful variation across countries, the overall picture is consistent—divorces across Europe tend to occur after many years of marriage, and long‑marriage divorces are far more widespread than early ones.

### Has divorce timing shifted over the last 30 Years?

Now that we’ve seen the overall European pattern and the geographic differences between countries, the next question is whether divorce timing has changed over the last three decades. Social norms, economic conditions, and family structures have all evolved since the mid‑1990s, and these shifts may influence when marriages are most likely to end. To explore this, we track how the share of divorces in each duration category has changed from 1995 to 2025.

```{r}
#| echo: false
# Prepare data for time trend
trend_data <- europe_divorce |>
  mutate(
    year = as.numeric(year),
    duration_std = factor(
      duration_std,
      levels = c("0-1", "1-2", "3-4", "5-9", "10+"))
  ) |>
  group_by(year, duration_std) |>
  summarise(
    mean_pct = mean(percent_of_total, na.rm = TRUE),
    .groups = "drop"
  )

# Plot: stacked area chart
ggplot(trend_data, aes(x = year, y = mean_pct, fill = duration_std)) +
  geom_area(alpha = 0.9) +
  scale_fill_manual(
    values = c(
      "0-1" = "#deebf7",  
      "1-2" = "#c6dbef",  
      "3-4" = "#9ecae1",   
      "5-9" = "#6baed6",   
      "10+" = "#3182bd"    
    )
  ) +
  labs(
    title = "How Divorce Timing Has Shifted Over Time",
    subtitle = "Average share of divorces in each duration category, 1995–2025",
    x = "Year",
    y = "Average Share of Divorces",
    fill = "Duration"
  ) +
  theme_minimal()


```

A clear pattern emerges from the time‑trend data. Early‑marriage divorces (0–4 years) remain consistently low throughout the entire period, with no indication that they are becoming more common. Mid‑marriage divorces in the 5–9‑year range form a stable middle layer, but the most striking feature of the chart is the persistent dominance of long‑marriage divorces. The 10+ category accounts for the largest share of divorces across all years. This stability over time suggests that structural and demographic factors—rather than a fixed psychological milestone—shape when divorces tend to occur.

### So does the Seven-Year Itch exist?

Taken together, the evidence shows that the “Seven‑Year Itch” is more cultural shorthand than statistical reality. Across Europe, divorces do not cluster around the seven‑year mark; instead, the typical divorce occurs well beyond it. Both the geographic patterns and the long‑term trends point to the same conclusion: European marriages tend to last many years before breaking down, with long‑marriage divorces forming the largest share across almost all countries. Mid‑marriage divorces (5–9 years) remain more common than early ones, but there is no sharp spike at seven years—only a steady progression from early to mid‑marriage durations and a long, stable dominance of the 10+‑year category. In this sense, the Seven‑Year Itch captures a general idea about relationships changing over time, but it does not correspond to any distinct moment in the data.

### References

United Nations Statistics Division (UNSD) 2026, Divorces by duration of marriage, UN Demographic Statistics Database, accessed 28 May 2026, <https://data.un.org/>.

Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R, Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E, Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K, Vaughan D, Wilke C, Woo K, Yutani H (2019). “Welcome to the tidyverse.” *Journal of Open Source Software*, *4*(43), 1686. doi:10.21105/joss.01686 <https://doi.org/10.21105/joss.01686>.

Wickham H (2025). *stringr: Simple, Consistent Wrappers for Common String Operations*. doi:10.32614/CRAN.package.stringr <https://doi.org/10.32614/CRAN.package.stringr>, R package version 1.6.0, <https://CRAN.R-project.org/package=stringr>.

Pebesma, E., & Bivand, R. (2023). Spatial Data Science: With Applications in R. Chapman and Hall/CRC. https://doi.org/10.1201/9780429459016

Massicotte P, South A (2026). *rnaturalearth: World Map Data from Natural Earth*. doi:10.32614/CRAN.package.rnaturalearth <https://doi.org/10.32614/CRAN.package.rnaturalearth>, R package version 1.2.0, <https://CRAN.R-project.org/package=rnaturalearth>.

South A, Michael S, Massicotte P (2024). *rnaturalearthdata: World Vector Map Data from Natural Earth Used in 'rnaturalearth'*. doi:10.32614/CRAN.package.rnaturalearthdata <https://doi.org/10.32614/CRAN.package.rnaturalearthdata>, R package version 1.0.0, <https://CRAN.R-project.org/package=rnaturalearthdata>.

Arel-Bundock V, Enevoldsen N, Yetman C (2018). “countrycode: An R package to convert country names and country codes.” *Journal of Open Source Software*, *3*(28), 848. <https://doi.org/10.21105/joss.00848>.

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### Unsexy Parts of the Workflow (Question 5)

A lot of the work behind this project was slow and unglamorous. One of the most tedious tasks was dealing with the completely unstandardised duration‑of‑marriage categories: countries used overlapping ranges, inconsistent labels, and sometimes open‑ended buckets, which meant I had to manually inspect and recode many variations before they could be grouped into five consistent bins. Early on, my cleaning steps even produced non‑unique country–year–duration rows, so I spent a lot of time re‑running the pipeline, checking intermediate tables, and making sure each country–year ended up with exactly one row per bin. I also had to harmonise country names, verify that joins didn’t silently drop rows, and repeatedly confirm that percentages summed correctly. None of this was technically difficult, but it was repetitive, manual, and essential for producing a dataset I could trust.

### Challenges Faced During the Analysis (Question 6) 
The biggest challenge was realising that my initial cleaning pipeline was structured incorrectly. I originally computed percentages before standardising duration categories, which produced duplicated rows and totals that didn’t add up. Fixing this required rethinking the entire workflow and moving the standardisation step earlier so that each country–year–duration combination was unique before aggregation. Another challenge was dealing with incomplete reporting across countries, which forced me to narrow the scope to Europe and make explicit assumptions—such as assigning a midpoint of 15 years to the open‑ended “10+” category. These choices forced me to make several non‑trivial assumptions and analytical decisions—especially around how to treat open‑ended categories and incomplete reporting—and working through them was essential for producing results that were both defensible and meaningful.

### Imperfect Parts and Future Improvements (Question 7) 
A key limitation of my analysis is that it ultimately focuses only on Europe, because many countries outside the region reported incomplete or inconsistent duration breakdowns, making global comparisons unreliable within the scope of this assignment. Another imperfect aspect is the way I had to group duration‑of‑marriage categories: the raw data uses overlapping ranges, inconsistent labels, and open‑ended bins, so forcing them into five standard categories required simplifying assumptions that inevitably blur some nuance. Although the cleaning pipeline now ensures unique country–year–duration rows, the underlying binning still reflects compromises rather than perfectly aligned definitions. In future work, I would like to expand the analysis beyond Europe by developing methods to handle missing or partial duration distributions, and refine the duration modelling by estimating continuous distributions or using probabilistic approaches for partially overlapping bins instead of relying on fixed midpoints and manual recoding.

### How I Used AI in This Assignment (Question 8)
I used AI mainly as a support tool for clarifying concepts, checking my reasoning, and improving the clarity of my written explanations. It was particularly helpful when I needed to articulate the logic behind my data‑cleaning steps or describe why certain transformations were necessary. However, AI was not consistently reliable when it came to coding: I often had to provide additional context, correct its assumptions about my data structure, or adapt its suggestions to fit the actual pipeline. In several cases—especially when my initial cleaning steps produced non‑unique country–year–duration rows—I questioned the AI’s output and verified it by inspecting intermediate tables myself. This process helped me understand the underlying issue more deeply and ultimately restructure the workflow. AI also helped me learn new ways to phrase methodological explanations, but I always validated the technical parts independently to ensure the analysis remained accurate and defensible.

### Iterative Process (Question 9) 
This is the link to my git repository <https://github.com/nwat0027/ETC5512-assignment4/>
To demonstrate my iterative process in completing the assignment, here are 4 commits on github that show changes of my analysis. Raw files are also provided in the "earlier_versions" folder.

#### Commit #b9046bc - "change the colour palette for better interpretation"
To improve the interpretability of the visualisation, I revised the colour palette used for the duration categories by using a sequential light‑to‑dark blue scale, where the lightest shade represents the shortest duration category and progressively darker shades represent longer durations. Raw
 
#### Commit #054466c - "edit logic of finding europe average"
Previously, I calculated the overall mean of percent_of_total across all years, which unintentionally compressed the values and didn’t reflect how divorce durations were distributed in a typical year. I realised this approach was misleading, so I switched to a two‑stage averaging process: first computing the average share of each duration bin within each year, and then averaging those yearly values across time. This affects the result and written analysis.

#### Commit #f7fb10c - "change steps in data cleaning" 
During data cleaning, I restructured the workflow so that the standardisation of duration‑of‑marriage categories happens before joining totals and computing percent_of_total. In the original pipeline, percentages were calculated first and only then collapsed into standardised bins, which caused duplicated rows, incorrect proportions, and totals that did not sum to one. By moving the standardisation step earlier, I ensured that each country–year–duration combination was unique before aggregation, allowing me to correctly sum values within bins and compute accurate percentages. This also impacted all the results and conclusions I drew, therefore the analysis was also rewritten.

#### Commit #ae36bf2 - "change calculation method from peak duration to weighted average for more meaningful map"
The previous map I created from the incorrectly processed data showed almost every country falling into the “10+” category, which made the visualisation uninformative and suggested that my earlier method was masking variation. Hence, instead of mapping the most common duration bin, I switched to calculating a weighted average duration for each country, which provides a more meaningful and interpretable summary of divorce timing across Europe.

:::
